Classless Java: Tuning Java Interfaces

### 2015-12-07 paper/sections/RelatedWork.tex
Related Wotk

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple
inheritance is very expressive, but difficult to model and implement, and can
cause difficulty (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about 
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid model
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; mixin model adds some
restrictions on the model; and trait model is the most restricted one (excluding
states, instantiation, etc).

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance by virtual
inheritance that deals with diamond problem by keeping only one copy of the base
class~\cite{ellis1990annotated}, however suffers from object initialization
problem as shown in~\cite{malayeri2009cz}. It bypasses all constructor calls to
virtual superclasses, which would potentially cause serious semantic errors. In
our approach, \mixin annotation processor has the full control over object
initialization, whose mechanism is transparent to users. Moreover, customized
factory methods are also allowed, if users are not satisfied with the default
generated \texttt{of} method.

\paragraph{Mixins.}
Mixins are a more restricted model than C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods do not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which aims to do
multiple inheritance without diamond problem. They divide inheritance into two
separate concepts: inheritance dependency and implementation inheritance. Using
a combination of \texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover, fields and
multiple inheritance can coexist. However as shown in the example in the
article, the untangling of inheritance also untangles class structure. In CZ,
not only the number of classes (from 4 to 6 in the stream example), but also the
class hierarchy complexity increases. \mixin does not complicate the
structure, and fields can also coexist with multiple inheritance.

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality, ignoring states and state
initialization. Classes, as object factories, require functionalities from
(multiple) traits. Java 8 interfaces are closely related to
traits: concrete method implementation are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavours of multiple inheritance in Java. Traits offer an algebra
of composition operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.  We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The author of ~\cite{bono14} agrees that our revised version is cleaner,
typesafe and more direct.

@@

There are also proposals for extending Java (before Java8) with traits. For
example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} extends
FJ~\cite{Igarashi01FJ} with statically-typed traits, adding trait-based
inheritance in Java.  Except for few, mostly syntactic details, their work can
be emulated with Java8 interfaces. There are also extensions to the original
trait model, which have operations that default methods and interfaces cannot
model, such as method renaming in \cite{reppy2006foundation}, which breaks
structural subtyping.

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with states is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
states, while still being quite expressive.  Moreover the ability to support
constructors adds additional expressivity, which is not available in approaches
such as Scala's traits/mixins.

@@

\subsection{ThisType/MyType/Extensibility}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namelly it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolish most advantages of interfaces.}

@@

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside of the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2)exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment. 
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. We are using covariant-return types to
simulate some uses of MyType. But our approach only works on method positive
positions, whereas MyType is more general, as it works on any
positions. Nevertheless our approach is still useful for modeling fluent
interfaces and solving expression problems,etc.
% The addition of MyType to a language will allow easy definition of
%binary methods, methods with recursive types (i.e., the same type of the
%receiver appears in the argument or return positions of methods), etc.

@@

\subsection{Meta-programming Competes with Language Extensions}
The most obvious solution to adding features to a language is language
extension. It is often implemented as syntactic extensions that can be desugared
to the base language. For example, the Scala compiler was extended to directly
supports XML syntax. However, this approach does not support combining multiple
extensions into one. Scala-Virtualized is an extension to Scala which allows to
blend shallow and deep embedding of DSLs. It redefines Scala structures to
method calls, which can be overridden by DSL implementors. Therefore
Scala-Virtualized also reinterprets syntax. These extensions are admittedly
useful, however they are essentially creating a fork of the language, and rarely
the new fork gains enough traction to become the main language release. % On this
% topic we mention
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language allowing
programmers to extend it with custom features by definitions in meta-DSLs (SDF,
Stratego, etc). Using SugarJ requires learning various DSLs while our approach
is just using Java. However its intention is different from ours: one of the
purposes of SugarJ is to allow easy defition of sugars while our approach aims to
implement the desired program transformation in pure Java.

On the other side, when the base language has a flexible enough syntax and
a fast and powerful enough reflection mechanism, we may just need to play with
operator overloading and other language tricks to discover that the language
feature we need can be expressed as a simple library in our language. For
example, consider SQLAlchemy in python.

Java-like languages tend to sit in the middle of two extremes: libraries
can not influence the type system, so many solutions valid in python or other dynamic
languages are not applicable, or have the cost of losing
safety.

Here compile time code generation comes at the rescue: if for a certain feature
(\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examines and enriches the code before compilation. No need to modify the
original source; for example we can work on temporary files. Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process: in this way
there is no need of temporary files, and there is a well-defined way of putting
multiple extensions together.

Other languages offer even stronger support to safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time and to generate code/classes that are transparently integrated in the program that is being generated/processed/compiled.
In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or unreasonable) would never generate ill-typed code. This is roughly equivalent to 
\textit{Self coherence}, that we manually proof in the Appendix.
Since MetaFjig is not working on annotated classes, there is no direct
equivalence on the overall theorem of the shown safety.

>>>>

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple
inheritance is very expressive, but difficult to model and implement, and can
cause difficulty (including the famous diamond (fork-join)
problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.) in reasoning about 
programs. To conciliate the need for expressive power and simplicity, many
models have been proposed, including C++ virtual inheritance,
mixins~\cite{bracha90mixin}, traits~\cite{scharli03traits}, and hybrid models
such as CZ~\cite{malayeri2009cz}.  They provide novel programming architecture
models in the OO paradigm. In terms of restrictions set on these models, C++
virtual inheritance aims at a relative general model; the mixin model adds some
restrictions; and the trait model is the most restricted one (excluding
states, instantiation, etc).

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance by
virtual inheritance, dealing with the diamond problem by keeping only
one copy of the base class~\cite{ellis1990annotated}. However C++
suffers from the object initialization problem~\cite{malayeri2009cz}.
It bypasses all constructor calls to virtual superclasses, which can
potentially cause serious semantic errors. In our approach, the \mixin
annotation has full control over object initialization,
and the mechanism is transparent to users. Moreover, customized factory
methods are also allowed: if users are not satisfied with the default
generated \texttt{of} method, they can implement their own.

\paragraph{Mixins.}
Mixins are a more restricted model than the C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, % they suffer from linearisation: the order of mixin application
% is relevant in often subtle and undesired ways. This hinders their usability
% and the ability of resolving conflicts:
the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in some cases and
restricts the flexibility of mixin composition. To fight this limitation, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raises the complexity, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods do not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

@@

\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which
aims to do multiple inheritance without the diamond problem. They
divide inheritance into two separate concepts: inheritance dependency
and implementation inheritance. Using a combination of
\texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover,
fields and multiple inheritance can coexist. However untangling
inheritance also untangles the class structure. In CZ, not only the
number of classes, but also the class hierarchy complexity
increases. \mixin does not complicate the class structure, and state
can also coexist with multiple inheritance.

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse and object factories. Traits, as units of reusable
code, contain only methods as reusable functionality, ignoring states and state
initialization. Classes, as object factories, require functionalities from
(multiple) traits. Java 8 interfaces are closely related to
traits: concrete method implementations are allowed (via the \textbf{default}
keyword) inside interfaces. The introduction of default methods opens the gate
for various flavours of multiple inheritance in Java. Traits offer an algebra
of composition operations like sum, alias and exclusion, providing explicit conflict
resolution. Former work~\cite{bono14} provides details on mimicking the trait
algebra through Java 8 interfaces.  We briefly recall the main points of their
encoding; however we propose a different representation of \textbf{exclusion}.
The first author of~\cite{bono14} agreed (via personal communication) that 
our revised version for exclusion is cleaner, typesafe and more direct.

@@

There are also proposals for extending Java with traits. For
example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} extends
FJ~\cite{Igarashi01FJ} with statically-typed traits, adding trait-based
inheritance in Java.  Except for few, mostly syntactic details, their work can
be emulated with Java 8 interfaces. There are also extensions to the original
trait model, which have operations that default methods and interfaces cannot
model, such as method renaming~\cite{reppy2006foundation}, which breaks
structural subtyping.

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with state is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
state, while still being quite expressive.  Moreover the ability to support
constructors adds expressivity, which is not available in approaches
such as Scala's traits/mixins.

@@

\subsection{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namelly it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolish most advantages of interfaces.}

@@

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in class based object oriented
programming, and is exasperated by the interface-based programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside of the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment. 
Admittedly, MyType greatly enhances the expressiveness and extensibility of
object-oriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But our approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Nevertheless, as illustrated with our examples and 
case studies, object interfaces are still very useful in many 
practical applications.
% The addition of MyType to a language will allow easy definition of
%binary methods, methods with recursive types (i.e., the same type of the
%receiver appears in the argument or return positions of methods), etc.

@@

\subsection{Meta-Programming Competes with Language Extensions}
The most obvious solution to adding features to a language is via syntactic
extensions. Syntactic extensions are often implemented as 
desugarings to the base language. For example,
the Scala compiler was extended to directly support XML syntax.
However, when syntactic extensions are independently created it is hard
to combine multiple extensions into one.
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language that
aims at making the creation and composition of syntactic sugar
extensions easy, by allowing programmers to extend Java with custom
features (typically for DSLs). However SugarJ's goals are different 
from language tuning: SugarJ aims at creating and composing new syntax; 
whereas language tuning merely reinterprets existing syntax. It is 
clear that reinterpreting existing syntax only can be limiting 
for some applications. However, when language tuning is possible 
it has the advantage that existing tools for the language work 
out-of-the-box (since the syntax is still the same); and composition
of independently developed tunings is straightforward.

Scala-Virtualized~\cite{}\bruno{reference!} is an extension to Scala, 
which allows blending shallow and deep embedding of DSLs. It redefines 
some of Scala's language constructs to method calls, 
which can be overridden by DSL implementors. Thus Scala Virtualized  
can also reinterpret syntax, and be seen as a form of language tuning. 
However, although many Scala's language constructs are supported, 
not all language constructs can be virtualized.

%These extensions are admittedly useful, however they are essentially 
%creating a fork of the language, and rarely the new fork gains enough 
%traction to become the main language release. % On this
% topic we mention

When the base language has a flexible enough syntax and a fast and
powerful enough reflection mechanism, we may just need to play with
operator overloading and other language tricks to discover that the
language feature we need can be expressed as a simple library in our
language. An example of this is SQLAlchemy\cite{}\bruno{reference} in
Python, which ...\bruno{does what? describe what SQLAlchemy is}
without requiring any syntactic extensions to Python.
Java-like languages tend to sit in the middle of two extremes:
libraries can not influence the type system, so many solutions valid
in Python or other dynamic languages are not applicable, or have the
cost of losing type-safety.

In Java-like languages compile time code generation comes at the rescue: if, for a certain feature
(\mixin in our case), it is possible to use the original language syntax to
\emph{express/describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examines and enriches the code before compilation. No need to modify the
original source (for example we can work on temporary files). 
Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process via annotation
processing. In this way
there is no need of temporary files, and there is a well-defined way of putting
multiple extensions together.

Other languages offer even stronger support for safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time. They generate code/classes that is transparently integrated in the program that is being generated/processed/compiled. In particular, MetaFjig offers a property called \emph{meta-level-soundness}. In short this property ensures by construction that library code (even if wrong or unreasonable) never generates ill-typed code. This is roughly equivalent to 
\textit{Self coherence}, that we have to manually prove.
Since MetaFjig is not working on annotated classes, there is no direct
equivalence on the overall theorem of the shown safety.\bruno{Don't understand 
the end of the sentence!}
